### 2021-07-08 rad-icfp-2021/purerad.tex
.

We consider a standard call-by-value language. 
It consists of a first-order functional language with arrays and a few typical second-order array operations. 
The types !T1,T2! and terms !e1,e2! are given in Figure~\ref{fig:source_grammar}.
We have included a minimal set of array operations for the sake of illustration,  it is not hard to add more.
See Section~\ref{sec:generalization}.

@@

The typing rules are given in Figure~\ref{fig:source_typesystem}. 
For scalar operations, we assume a given set of operations, including $+$ and $*$. 
!op1! and !op2! denote respectively a unary and a binary operation on reals. 
These operations represent smooth total functions, 
but again this can be easily generalized (\S\ref{sec:generalization}).  
Typical examples include !cos, exp, +, *!. 
We use infix notation for binary operators.

@@

\input{figures/source_type_system}

@@

The types and terms are presented in Figure~\ref{fig:target_grammar}.
Our lambda abstractions take $n$ arguments as we are not interested here with partial applications. 
In fact, the lambda abstractions introduced by reverse-mode will be removed during partial evaluation, 
and the notation with lambda abstractions having $n$ bound variables makes reading slightly easier.
We note that we don't actually need the full power of higher-order because we only use lambda abstractions over variables of ground types
and let expressions binding such lambda abstractions. We only need the target language to be second-order.
The type system for the extended grammar of the target language is presented in Figure~\ref{fig:target_typesystem}.

@@

\input{figures/target_type_system}

>>>>

We consider a standard call-by-value language. 
It consists of a first-order functional language with arrays and a few typical second-order array operations. 
The types !T1,T2!, terms !e1,e2!, and typing rules are given in Figure~\ref{fig:source_grammar}.
We have included a minimal set of array operations for the sake of illustration,  it is not hard to add more.
See Section~\ref{sec:generalization}.

@@

For scalar operations, we assume a given set of operations, including $+$ and $*$. 
!op1! and !op2! denote respectively a unary and a binary operation on reals. 
These operations represent smooth total functions, 
but again this can be easily generalized (\S\ref{sec:generalization}).  
Typical examples include !cos, exp, +, *!. 
We use infix notation for binary operators.

@@

@@

The grammar for types and terms along with the type system of the target language are presented in Figure~\ref{fig:target_grammar}.
Our lambda abstractions take $n$ arguments as we are not interested here with partial applications. 
In fact, the lambda abstractions introduced by reverse-mode will be removed during partial evaluation, 
and the notation with lambda abstractions having $n$ bound variables makes reading slightly easier.
We note that we don't actually need the full power of higher-order because we only use lambda abstractions over variables of ground types
and let expressions binding such lambda abstractions. We only need the target language to be second-order.

@@

### 2021-07-08 rad-icfp-2021/unf.tex
.

A judgment in Source UNF is a triple !(T$_1$,e,T$_2$)! where !T$_1$! and !T$_2$! are lists of types of Source, and !e! is an expression of the Source UNF.
The typing rules are detailed in Figure~\ref{fig:source_unf_typesystem}.
$\reals^{\times n}$ is a notation for $\reals,\ldots,\reals$ with $n$ factors $\reals$. 
When !T$_1$! and !T$_2$! are lists of types and !A! a type, the operator !,! denotes snoc in !T$_1$,A! and denotes append in !T$_1$,T$_2$!.

\input{figures/source_unf_typesystem}

@@

The Target UNF is an extension of Source UNF. 
Its types are lists of types of Target. 
Higher-order types are needed for the continuation introduced by reverse-mode. 
In addition, it contains several new primitives which represent the transpose Jacobian of the primitives of Source UNF.
Given a type !T$\defeq$ [A$_1$,$\ldots$,A$_n$]!, we write \underline{T} for the type !A$_1$ $\times\ldots\times$ A$_n$! of Target. 
Target UNF has an internal composition $\icomp$. 
!e$_{1}$ $\icomp$ e$_{2}$! represents precomposition of a curried function !e$_{1}$! by a term !e$_{2}$!.
It is used by the reverse-mode transformation to pre-compose the continuation term by a transpose Jacobian.
The grammar of Target UNF is given in Figure~\ref{fig:unf_target_grammar} and its type system in Figure~\ref{fig:target_unf_typesystem}.

@@

\input{figures/target_unf_typesystem}

>>>>

A judgment in Source UNF is a triple !(T$_1$,e,T$_2$)! where !T$_1$! and !T$_2$! are lists of types of Source, and !e! is an expression of the Source UNF.
The typing rules are detailed in Figure~\ref{fig:unf_source_grammar}.
$\reals^{\times n}$ is a notation for $\reals,\ldots,\reals$ with $n$ factors $\reals$. 
When !T$_1$! and !T$_2$! are lists of types and !A! a type, the operator !,! denotes snoc in !T$_1$,A! and denotes append in !T$_1$,T$_2$!.

@@

The Target UNF is an extension of Source UNF. 
Its types are lists of types of Target. 
Higher-order types are needed for the continuation introduced by reverse-mode. 
In addition, it contains several new primitives which represent the transpose Jacobian of the primitives of Source UNF.
Given a type !T$\defeq$ [A$_1$,$\ldots$,A$_n$]!, we write \underline{T} for the type !A$_1$ $\times\ldots\times$ A$_n$! of Target. 
Target UNF has an internal composition $\icomp$. 
!e$_{1}$ $\icomp$ e$_{2}$! represents precomposition of a curried function !e$_{1}$! by a term !e$_{2}$!.
It is used by the reverse-mode transformation to pre-compose the continuation term by a transpose Jacobian.
The grammar and type system of target UNF is given in Figure~\ref{fig:unf_target_grammar}.

@@
